Conformal Counterfactual Explanations

### 2023-03-22 paper/paper.tex
intro and background done

Finding ways to generate \textit{plausible} counterfactuals has been one of the primary concerns. To this end, \citet{joshi2019realistic} were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $x\sim\mathcal{X}$. To learn the latent embedding, they introduce a surrogate model. In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $x^*\sim \mathcal{G}(z)$ where $\mathcal{G}$ denotes the decoder part of the VAE. Provided the surrogate model is well-trained, their proposed approach can yield compelling counterfactual explanations like the one in the centre panel of Figure~ref{fig:vae}.

@@

Note that Definition~\ref{def:plausible} subsumes the notion of plausible counterfactual paths, since we can simply apply it to each counterfactual state along the path.

@@

>>>>

Finding ways to generate \textit{plausible} counterfactuals has been one of the primary concerns. To this end, \citet{joshi2019realistic} were among the first to suggest that instead of searching counterfactuals in the feature space $\mathcal{X}$, we can instead traverse a latent embedding $\mathcal{Z}$ that implicitly codifies the data generating process (DGP) of $x\sim\mathcal{X}$. To learn the latent embedding, they introduce a surrogate model. In particular, they propose to use the latent embedding of a Variational Autoencoder (VAE) trained to generate samples $x^* \leftarrow \mathcal{G}(z)$ where $\mathcal{G}$ denotes the decoder part of the VAE. Provided the surrogate model is well-trained, their proposed approach ---REVISE--- can yield compelling counterfactual explanations like the one in the centre panel of Figure~\ref{fig:vae}.

@@

Note that Definition~\ref{def:plausible} is consistent with the notion of plausible counterfactual paths, since we can simply apply it to each counterfactual state along the path.

Surrogate models offer an obvious solution to achieve this objective. Unfortunately, surrogates also introduce a dependency: the generated explanations no longer depend exclusively on the Black Box Model itself, but also on the surrogate model. This is not necessarily problematic if the primary objective is not to explain the behaviour of the model but to offer recourse to individuals affected by it. It may become problematic even in this context if the dependency turns into a vulnerability. To illustrate this point, we have used REVISE \citep{joshi2019realistic} with an underfitted VAE to generate the counterfactual in the right panel of Figure~\ref{fig:vae}: in this case, the decoder step of the VAE fails to yield plausible values ($\{x^{\prime} \leftarrow \mathcal{G}(z)\} \not\sim \mathcal{X}|t$) and hence the counterfactual search in the learned latent space is doomed.

@@

\section{A Framework for Conformal Counterfactual Explanations}\label{cce}
